---
title: "Differential Gene Expression Analysis - Blue Catfish Testis Tissue Samples"
output:
  html_document:
    df_print: paged
---

Aging is associated with widespread changes in gene expression across tissues. Reproductive tissues are particularly sensitive to these shifts given their central role in fertility and gonadal function. In this study, RNA-sequencing was used to compare gene expression profiles in testis tissue from blue catfish (Ictalurus furcatus) at two age points (2 years and 7 years) to identify genes whose expression changes significantly with age and sexual maturity.

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/samit/Desktop/class_assignment_2")
```

```{r}
# Load edgeR
library(edgeR)
```

```{r}
# Read count data
count_data <- read.csv("Counts.csv", header = TRUE, row.names = 1)
```

## Step 1: Differential Expression Analysis (edgeR)

```{r}
# Annotate the gene IDs with their common gene names
library(rtracklayer)
annotation <- import("bluecatfish_annotation.gtf")
annotation_df <- as.data.frame(annotation)
```

```{r}
# Build a gene_id which is a geneName lookup table from the annotation
gene_lookup <- unique(annotation_df[!is.na(annotation_df$transcript_id),
                                     c("transcript_id", "gene")])
names(gene_lookup) <- c("gene_id", "GeneName")
```

```{r}
# Define the treatments / columns 1-3 = Age2, columns 4-6 = Age7
Condition <- factor(rep(c("Age2", "Age7"), each = 3), levels = c("Age2", "Age7"))
```

```{r}
# Build DGEList
dge <- DGEList(counts = count_data, group = Condition)
```

```{r}
# Filter low count genes
keep <- filterByExpr(dge)
dge <- dge[keep, , keep.lib.sizes = FALSE]
```

```{r}
# TMM normalization / normalize the gene counts
dge <- calcNormFactors(dge, method = "TMM")
```

```{r}
# Design matrix
design <- model.matrix(~0 + Condition)
colnames(design) <- levels(Condition)
```

```{r}
# Estimate dispersion and fit QL model
dge <- estimateDisp(dge, design)
fit <- glmQLFit(dge, design)
```

```{r}
# Comparison - Age7 vs Age2
contrast <- makeContrasts(Age7 - Age2, levels = design)
qlf <- glmQLFTest(fit, contrast = contrast)
```

```{r}
# Get results, filter by FDR < 0.01
results <- topTags(qlf, n = Inf)$table
results$gene_id <- rownames(results)
results <- merge(gene_lookup, results, by = "gene_id", all.y = TRUE)
sig <- results[results$FDR < 0.01, ]
```

```{r}
# Split into up/down regulated (>= 3-fold change)
fold_change_cutoff <- 3
up <- sig[sig$logFC > log2(fold_change_cutoff), ]
down <- sig[sig$logFC < -log2(fold_change_cutoff), ]
```

```{r}
# to get the gene count
gene_counts <- data.frame(Category = c("Upregulated", "Downregulated"),
                          Count = c(nrow(up), nrow(down)))
print(gene_counts)
```

```{r}
# Save results
write.csv(results, "AllResults_Age7_vs_Age2.csv")
write.csv(up, "Upregulated_Age7_vs_Age2.csv")
write.csv(down, "Downregulated_Age7_vs_Age2.csv")
```

## Step 2: PCA and Sample Clustering

```{r}
library(ggplot2)
library(pheatmap)
```

```{r}
# Creating the PCA plot
logCPM <- cpm(dge, log = TRUE)
pca <- prcomp(t(logCPM), scale. = TRUE)
percentVar <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 1)
pca_df <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2], Condition = Condition, Sample = colnames(logCPM))
```

```{r}
p_pca <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Condition)) +
  geom_point(size = 4) +
  geom_text(aes(label = Sample), vjust = -1, size = 3.5, show.legend = FALSE) +
  scale_color_manual(values = c("Age2" = "orangered", "Age7" = "navy")) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_minimal()
p_pca
ggsave(filename = "PCA_Plot_Age2_vs_Age7.png", plot = p_pca, width = 6, height = 5, dpi = 600)
```

```{r}
#Sample clustering heat map generation
sample_cor <- cor(logCPM)
annotation_col <- data.frame(Condition = Condition, row.names = colnames(logCPM))
pheatmap(sample_cor, annotation_col = annotation_col)
pheatmap(sample_cor, annotation_col = annotation_col, filename = "Heatmap_SampleCorrelation_Age2_vs_Age7.png")
```

## Step 3: DEG Visualization

```{r}
# Creating the volcano plot
results_volcano <- topTags(qlf, n = Inf)$table
results_volcano$Category <- "Not Significant"
results_volcano$Category[results_volcano$FDR < 0.01 & results_volcano$logFC > log2(fold_change_cutoff)] <- "Upregulated"
results_volcano$Category[results_volcano$FDR < 0.01 & results_volcano$logFC < -log2(fold_change_cutoff)] <- "Downregulated"
up_count <- sum(results_volcano$Category == "Upregulated")
down_count <- sum(results_volcano$Category == "Downregulated")
```

```{r}
# Volcano plot
ggplot(results_volcano, aes(logFC, -log10(FDR), color = Category)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_color_manual(values = c(Upregulated = "orangered", Downregulated = "navy", "Not Significant" = "grey")) +
  geom_vline(xintercept = c(-log2(fold_change_cutoff), log2(fold_change_cutoff)), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.01), linetype = "dashed") +
  annotate("text", x = Inf, y = Inf, label = paste("Up:", up_count), hjust = 1.1, vjust = 2, color = "orangered", fontface = "bold") +
  annotate("text", x = -Inf, y = Inf, label = paste("Down:", down_count), hjust = -0.1, vjust = 2, color = "navy", fontface = "bold") +
  theme_minimal()
ggsave("Volcano_Age7_vs_Age2.png", width = 6, height = 6, dpi = 300)
```

```{r}
# Significant DEG heat map
library(dplyr)
# Get the top 20 most significant genes by FDR, excluding unnamed "LOC..." genes
top20_genes <- sig %>%
  filter(!startsWith(GeneName, "LOC")) %>%
  arrange(FDR) %>%
  pull(gene_id) %>%
  head(20)
```

```{r}
top20_expr <- logCPM[rownames(logCPM) %in% top20_genes, ]
# Use gene names as row labels where available, gene_id otherwise
gene_labels <- gene_lookup$GeneName[match(rownames(top20_expr), gene_lookup$gene_id)]
gene_labels[is.na(gene_labels)] <- rownames(top20_expr)[is.na(gene_labels)]
rownames(top20_expr) <- gene_labels
```

```{r}
pheatmap(top20_expr,
         cluster_rows = TRUE,
         show_rownames = TRUE,
         border_color = NA,
         fontsize = 10,
         scale = "row",
         fontsize_row = 10)
pheatmap(top20_expr,
         cluster_rows = TRUE,
         show_rownames = TRUE,
         border_color = NA,
         fontsize = 10,
         scale = "row",
         fontsize_row = 10,
         filename = "Top20_SigGenes_Heatmap_Age7_vs_Age2.png")
```

## Step 4: Gene Ontology (GO) Enrichment Analysis

```{r}
# Gene ontology analysis
library(gprofiler2)
# Use gene symbols against zebrafish annotation, exclude LOC-prefixed and missing names
up_names <- up$GeneName[!is.na(up$GeneName) & !startsWith(up$GeneName, "LOC")]
down_names <- down$GeneName[!is.na(down$GeneName) & !startsWith(down$GeneName, "LOC")]
```

```{r}
# GO enrichment
go_up <- gost(query = up_names, organism = "drerio", sources = "GO")
go_down <- gost(query = down_names, organism = "drerio", sources = "GO")
go_up$result
go_down$result
```

```{r}
# Save results
write.csv(go_up$result[, 1:13], "GO_Upregulated_Age7_vs_Age2.csv", row.names = FALSE)
write.csv(go_down$result[, 1:13], "GO_Downregulated_Age7_vs_Age2.csv", row.names = FALSE)
```

```{r}
#Draw the GO bar plot for upregulated GO term at age 7
go_colors <- c("GO:BP" = "steelblue", "GO:CC" = "darkorange", "GO:MF" = "forestgreen")
# Top 10 GO terms by significance - Upregulated
top_go_up <- go_up$result %>% arrange(p_value) %>% head(10)
ggplot(top_go_up, aes(x = reorder(term_name, -log10(p_value)), y = -log10(p_value), fill = source)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = go_colors) +
  labs(x = NULL, y = "-log10(p-value)", fill = "GO Category", title = "Top GO Terms - Upregulated") +
  theme_minimal()
ggsave("GO_Upregulated_BarPlot.png", width = 8, height = 6, dpi = 300)
```
```{r}
#Draw the GO bar plot for downregulated GO term at age 7
go_colors <- c("GO:BP" = "steelblue", "GO:CC" = "darkorange", "GO:MF" = "forestgreen")
# Top 10 GO terms by significance - Downregulated
top_go_down <- go_down$result %>% arrange(p_value) %>% head(10)
ggplot(top_go_down, aes(x = reorder(term_name, -log10(p_value)), y = -log10(p_value), fill = source)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = go_colors) +
  labs(x = NULL, y = "-log10(p-value)", fill = "GO Category", title = "Top GO Terms - Downregulated") +
  theme_minimal()
ggsave("GO_Downregulated_BarPlot.png", width = 8, height = 6, dpi = 300)
```

